~\Large Eras in baseball: Change-Ups and Change Points: An Exploration of Baseball's Historic Eras

### 2023-07-19 manuscript/paper.Rmd
updated.

---
title: |
  |  \Large Eras in baseball: Change-Ups and Change Points: An Exploration of Baseball's Historic Eras
author: | 
  | \large Mena Whalen \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Center for Data Science and Consulting \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`mwhalen3@luc.edu`](mailto:mwhalen3@luc.edu) \vspace{-1mm}
  |
  | \large Brian M. Mills \vspace{-1.1mm}
  | \normalsize College of Education \vspace{-1mm}
  | \normalsize University of Texas at Austin \vspace{-1mm}
  | \normalsize Austin, TX 60660 \vspace{-1mm}
  | \normalsize [`brian.mills@austin.utexas.edu`](mailto:brian.mills@austin.utexas.edu) \vspace{-1mm}
  |
  | \large Gregory J. Matthews \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Center for Data Science and Consulting \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}
abstract: |
  | Baseball is some weird and wild shit. \vspace{2mm}
  | *Keywords*: change point analysis, baseball, 
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

>>>>

---
title: |
  |  \Large Eras in baseball: Change-Ups and Change Points: An Exploration of Baseball's Historic Eras
author: | 
  | \large Mena Whalen \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Center for Data Science and Consulting \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`mwhalen3@luc.edu`](mailto:mwhalen3@luc.edu) \vspace{-1mm}
  |
  | \large Brian M. Mills \vspace{-1.1mm}
  | \normalsize College of Education \vspace{-1mm}
  | \normalsize University of Texas at Austin \vspace{-1mm}
  | \normalsize Austin, TX 78712 \vspace{-1mm}
  | \normalsize [`brian.mills@austin.utexas.edu`](mailto:brian.mills@austin.utexas.edu) \vspace{-1mm}
  |
  | \large Gregory J. Matthews \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Center for Data Science and Consulting \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}
abstract: |
  | Baseball is some weird and wild shit. \vspace{2mm}
  | *Keywords*: change point analysis, baseball, 
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

### 2023-07-19 manuscript/paper.tex
updated.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace} \setstretch{1.15} \usepackage{float} \floatplacement{figure}{t}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  colorlinks=true,
  linkcolor={cyan},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={cyan},
  pdfcreator={LaTeX via pandoc}}

\title{~\Large Eras in baseball: Change-Ups and Change Points: An
Exploration of Baseball's Historic Eras}
\author{\large Mena Whalen \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
\normalsize Center for Data Science and Consulting \vspace{-1mm}\\
\normalsize Loyola University Chicago \vspace{-1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1mm}\\
\normalsize \href{mailto:mwhalen3@luc.edu}{\texttt{mwhalen3@luc.edu}}
\vspace{-1mm}\\
\strut \\
\large Brian M. Mills \vspace{-1.1mm}\\
\normalsize College of Education \vspace{-1mm}\\
\normalsize University of Texas at Austin \vspace{-1mm}\\
\normalsize Austin, TX 60660 \vspace{-1mm}\\
\normalsize \href{mailto:brian.mills@austin.utexas.edu}{\texttt{brian.mills@austin.utexas.edu}}
\vspace{-1mm}\\
\strut \\
\large Gregory J. Matthews \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
\normalsize Center for Data Science and Consulting \vspace{-1mm}\\
\normalsize Loyola University Chicago \vspace{-1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1mm}\\
\normalsize \href{mailto:gmatthews1@luc.edu}{\texttt{gmatthews1@luc.edu}}
\vspace{-1mm}}
\date{}

@@

\hypertarget{data}{%
\subsection{Data}\label{data}}

We obtained multi-year baseball statistics from GREG??. Over the course
of the sport's history, numerous statistics have been collected, and for
our analysis, we focused on year-end statistics pertaining to the teams.
To ensure consistency in our methodology, it was necessary for each team
to have existed from 1900 until the end of the dataset in 2020. This
timeframe of 120 years, denoted as T, encompassed a total of 16
franchise teams, namely NYY, BOS, LAD, ATL, CHW, CHC, CIN, CLE, DET,
BAL, SFG, OAK, PHI, PIT, STL, and MIN. The statistics of interest for
these teams include runs, hits, home runs, balls, strikeouts, at-bats,
stolen bases, number of games played in a season, attendance, runs
against, hits against, home runs against, balls against, and strikeouts
against. However, a few statistics, such as balls, stolen bases,
strikeouts, and attendance, had missing data for certain years. To
address this, we employed multiple imputations using classification and
regression trees (cite mice/cart) to fill in the missing values and
ensure consistent data across all analyses.

\hypertarget{eras-and-dynasties}{%
\subsection{Eras and Dynasties}\label{eras-and-dynasties}}

\hypertarget{eras}{%
\subsubsection{Eras}\label{eras}}

Since we are interested in finding where in time these different eras of
baseball have occurred, we first examine each baseball statistic of
interest as their own analysis for all 16 teams. Those baseball
statistics are home runs, strikeouts, balls, stolen bases, runs, and
attendance which were all standardized to the number of games played in
a season since we are interested in seeing how the different statistics
change from season to season. Using the methodology from Cho and Fry
(CITE) a thresholding parameter was determined from bootstrapping and
then used to find where change points in mean and variance existed in
the panel of teams data of a given baseball statistic. To examine the
idea of changes throughout teams and types of statistics (initially),
for each season the average baseball statistic was found from all the
teams that year and then an average time series was created of each of
the types of statistics of interest. Then the change point methodology
was performed on those 6 time series of average statistic types.

\#\#\#Dynasties

Continuing our investigation, we explored the concept of ``dynasties''
within modern baseball teams, irrespective of their length of a teams
existence. A dynasty, in this context, refers to a period characterized
by significant success or potentially hardships. Such a dynasty is
identified when a team experiences a noticeable change across all their
statistics in a given season, whether it be in a positive or negative
direction. This approach aligns well with our methodology, as it
involves examining panels of time series data encompassing all
statistics for a singular team. The statistics of interest for this
analysis include runs, hits, home runs, balls, strikeouts, runs against,
hits against, home runs against, balls against, and strikeouts against.
To ensure comparability, these statistics were standardized using the
season average and standard deviation derived from all teams present in
each respective season. By standardizing the statistics, we capture a
team's offensive and defensive performance. Notably, if a team exhibits
a high and positive offensive performance, it can be balanced to zero by
large and negative defensive statistics when standardized. Therefore, if
a team excels in both aspects of the game, it would have numerous time
points above zero, whereas underperforming teams would display many
below-zero values.For each team, we examined their ten statistics,
analyzing both mean and variance changes to identify potential change
points. By scrutinizing these indicators, we aimed to uncover shifts in
a team's performance and assess their significance within the context of
dynasties.

@@

\hypertarget{tables}{%
\subsection{Tables}\label{tables}}

\begin{longtable}[]{@{}lll@{}}
\toprule\noalign{}
Mean/Variance CPT & Franchise & Change Point Year \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
mean & ANA & 1977 \\
mean & ATL & 1900, 1990 \\
mean & BAL & 1959, 1999 \\
mean & BOS & 1918, 1937 \\
mean & CHC & 1891 \\
mean & CHW & 1970 \\
mean & CIN & 1918, 1952 \\
mean & CLE & 1944, 1959, 1993 \\
mean & DET & 1958, 1988 \\
mean & FLA & 2011 \\
mean & HOU & 1999 \\
mean & KCR & 1997 \\
mean & LAD & 1940, 1961 \\
mean & MIL & 1977, 1996 \\
mean & MIN & 1959 \\
mean & PHI & 1917, 1949 \\
mean & PIT & 1939 \\
mean & SDP & 1977 \\
mean & SEA & 1986, 1999 \\
mean & SFG & 1903, 1973 \\
mean & STL & 1937 \\
mean & TBD & 2007 \\
mean & TEX & 1973, 1990 \\
mean & WSN & 2011 \\
variance & ATL & 1887 \\
variance & BOS & 2008 \\
variance & CHW & 1969 \\
variance & CLE & 1966, 1981, 1993 \\
variance & DET & 1942 \\
variance & HOU & 2010 \\
variance & NYM & 1974 \\
variance & PIT & 1902, 2001 \\
variance & SDP & 1977 \\
variance & SFG & 1918 \\
\end{longtable}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2889}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5111}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Mean/Variance CPT
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Baseball Statistic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Change Point Year
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
mean & att\_per\_game & 1945, 1975, 1992 \\
mean & bb\_per\_game & 1927 \\
mean & hr\_per\_game & 1920, 1946, 1967, 1993 \\
mean & r\_per\_game & 1919, 1940 \\
mean & sb\_per\_game & 1919 \\
mean & so\_per\_game & 1956, 1993, 2009 \\
variance & att\_per\_game & 1944, 2008 \\
variance & hr\_per\_game & 1927, 1944, 1995, 2007 \\
variance & r\_per\_game & 1935 \\
variance & sb\_per\_game & 1919, 1966 \\
variance & so\_per\_game & 1990 \\
\end{longtable}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace} \setstretch{1.15} \usepackage{float} \floatplacement{figure}{t}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  colorlinks=true,
  linkcolor={cyan},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={cyan},
  pdfcreator={LaTeX via pandoc}}

\title{~\Large Eras in baseball: Change-Ups and Change Points: An
Exploration of Baseball's Historic Eras}
\author{\large Mena Whalen \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
\normalsize Center for Data Science and Consulting \vspace{-1mm}\\
\normalsize Loyola University Chicago \vspace{-1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1mm}\\
\normalsize \href{mailto:mwhalen3@luc.edu}{\texttt{mwhalen3@luc.edu}}
\vspace{-1mm}\\
\strut \\
\large Brian M. Mills \vspace{-1.1mm}\\
\normalsize College of Education \vspace{-1mm}\\
\normalsize University of Texas at Austin \vspace{-1mm}\\
\normalsize Austin, TX 78712 \vspace{-1mm}\\
\normalsize \href{mailto:brian.mills@austin.utexas.edu}{\texttt{brian.mills@austin.utexas.edu}}
\vspace{-1mm}\\
\strut \\
\large Gregory J. Matthews \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
\normalsize Center for Data Science and Consulting \vspace{-1mm}\\
\normalsize Loyola University Chicago \vspace{-1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1mm}\\
\normalsize \href{mailto:gmatthews1@luc.edu}{\texttt{gmatthews1@luc.edu}}
\vspace{-1mm}}
\date{}

@@

@@

\begin{verbatim}
## [1]  20  46  67  94 106
\end{verbatim}

\begin{verbatim}
##  b 
## 48
\end{verbatim}

\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-1-1} \end{center}

\hypertarget{results-1}{%
\subsection{Results}\label{results-1}}

\begin{longtable}[]{@{}lrr@{}}
\toprule()
franch\_id & threshold & years \\
\midrule()
\endhead
NYY & 4.801818 & 1918 \\
BOS & 4.427752 & 1917 \\
BOS & 4.427752 & 1929 \\
BOS & 4.427752 & 1948 \\
BOS & 4.427752 & 1994 \\
LAD & 4.776867 & 1919 \\
LAD & 4.776867 & 1955 \\
ATL & 5.393778 & 1918 \\
ATL & 5.393778 & 1949 \\
CHW & 4.326192 & 1918 \\
CHC & 4.277577 & 1919 \\
CHC & 4.277577 & 1940 \\
CIN & 4.249476 & 1929 \\
CIN & 4.249476 & 1952 \\
CLE & 4.405339 & 1920 \\
CLE & 4.405339 & 1939 \\
DET & 3.926054 & 1927 \\
DET & 3.926054 & 1940 \\
BAL & 4.641706 & 1919 \\
BAL & 4.641706 & 1942 \\
SFG & 4.319634 & 1919 \\
SFG & 4.319634 & 1937 \\
OAK & 3.819294 & 1920 \\
OAK & 3.819294 & 1936 \\
OAK & 3.819294 & 1951 \\
PHI & 4.123104 & 1918 \\
PHI & 4.123104 & 1937 \\
PIT & 4.175793 & 1920 \\
PIT & 4.175793 & 1946 \\
STL & 5.444239 & 1919 \\
STL & 5.444239 & 1941 \\
MIN & 4.498666 & 1919 \\
MIN & 4.498666 & 1942 \\
\bottomrule()
\end{longtable}
